---
title: "quack_session3_2026"
author: "Nina Schoener"
date: "2026-09-24"
output: html_document
---

[Modeled after QuACK 2025 Summer Session 2 by Victoria Keating, Sophie Regan, and Nina Schoener]

## Loading our packages

```{r}

# let's start by installing and loading tidyverse in full using the method we went over

#install.packages("tidyverse")


# tidylog is a useful add-on that will give you feedback on what your tidyverse commands have done. let's add it! 

#install.packages("tidylog")


# you may get a warning saying "the following objects are masked from..."
# this means that there are some functions in the package you just loaded with the same name as functions in a different package (or base R)
# it's ok -- just be careful about this! if you want to be extra sure you're using the right function, you can always type "packagename::" before it!
# pro-tip: this also lets you use a particular function from a package without having to load it into your library first!

```

## Reading in our data

For today's session, we have collected (read: invented) some data about people's food choices and whether they are getting enough servings of fruits and vegetables on a daily basis. We're going to use this data to demonstrate some helpful data reshaping and merging functions. The csv files called "fruit_vegetable_servings.csv" and "participant_demographics.csv" contain the new data you will need. 


```{r}

# let's start by loading in both datasets using read_csv



# let's get familiar with the veggies data first
# what function should we call to see what kind of columns are in this dataframe?



# this dataframe is telling us how many servings of fruits and vegetables each participant had with each meal (presumably throughout a day). check the column types to make sure they're correct. the names are kind of annoying but you will see shortly that they will actually come in handy. 

# but there isn't a whole lot to work with in terms of participant-level variables
# let's see what the demo dataframe has to offer



# great, this dataframe offers us info on age, gender, location, and favorite cuisine!
# is there anything we need to clean up about this dataframe before continuing?



```



### Joining our dataframes

If we want to know anything about how participant-level variables are associated with veggie intake, we're going to need to put these two dataframes together. This is called joining two dataframes. 

```{r}

# there are several ways we can combine these dfs, but let's first use inner_join()	
# an inner join will keep ONLY the rows that exist in BOTH dataframes
# in order to identify these rows, you need to have one column in common across the two dataframes
# what column is that in our case?




# let's try another commonly used join method




# take a moment to choose another join method from below and try it on our two dataframes



# INNER JOIN: Keep rows where id exists in BOTH df1 and df2

# LEFT JOIN: Keep ALL rows from df1, add matching rows from df2, NA if no match

# RIGHT JOIN: Keep ALL rows from df2, add matching rows from df1, NA if no match

# FULL JOIN: Keep ALL rows from BOTH dataframes, NA if no match in one

# ANTI JOIN: Return rows from df1 that DO NOT have a match in df2

# SEMI JOIN: Return rows from df1 that DO have a match in df2 (but only columns from df1)
```

## Pivoting our data

Let's say we want to know how many servings of fruits and vegetables a day people eat by favorite cuisine. In order to do this, we're going to pivot the dataframe so that we can easily find the average number of servings of fruits and vegetables per day by participant, by meal, by cuisine, etc. 


```{r}

# we're going to start by pivoting the inner_joined data long. we're using this one because the left_joined data has NAs in it that we don't want to include. 
# let's use pivot() to reshape our data into long form
# you'll need to specify which columns you want to pivot, what you want the new variable for your column names to be called, and what you want the new variable for your column values to be called
# you can optionally specify some other things -- run ?pivot_longer to learn more



# great, now let's summarize this data by participant and by meal



# let's quickly plot this data -- you'll see why it was useful to pivot!




# pivoting made it so that all of the servings were in the Servings column, so we could summarize (sum, mean, etc) just that one column for our y-axis, and use the Meal and Participant columns to break up the data into columns in the plot


# let's practice doing this again, this time using data about the individuals in the dataset, rather than just their PID and the meal type: let's see if their favorite cuisine is associated with the amount of veggies they eat



# let's try plotting the data again



```


### Saving your long dataframe to a csv file


```{r}
# let's save our long dataframe to a new csv file in case we need it again in the future!



```

