---
title: "Fall 2025 Quack Session 8"
author: "Victoria Keating & Sophie Regan"
date: "2025-07-23"
output: html_document
---

# Loading
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```


# Warm Up Activity
```{r}
# Load in the data
GradLife <- 

# Number of respondents in the dataset


#mean, median, range


# Just for Cognition and Social, compare first caffeine usage
GradLife_CogSoc<-

# plot it


```


# T-Test
```{r}
# View average self-care by area


# Filter only clinical and developmental lab members


# Check how many people per lab


# Run the t-test

```




# Correlations
```{r}
# Plot the relationship

# Correlation test

```
# ANOVA
When you want to compare means across more than two groups, a t-test isn't enough — that's where ANOVA (Analysis of Variance) comes in.

ANOVA checks if there is any significant difference in means across the groups. It doesn't tell you which groups differ — just whether at least one group is different. You can follow up with post-hoc tests if needed.
```{r}
# Check how many areas are in the dataset


# Fit ANOVA model


# Plot the means with boxplots


# Let's do a post-hoc on the ANOVA




```

# Linear Regression
```{r}
# Run the regression

```

# Logistic Regression
```{r}

# Predict probabilities


# Plot

```




