---
title: "Fall 2025 Quack Session 8"
author: "Victoria Keating & Sophie Regan"
date: "2025-07-23"
output: html_document
---

# Loading
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```


# Warm Up Activity
```{r}
# Load in the data
GradLife <- read.csv("PsychGradLife_SemWide8.csv")

# Number of respondents in the dataset
n<-nrow(GradLife)

#mean, median, range
mean(GradLife$selfcare_score_S3)

median(GradLife$selfcare_score_S3)

range(GradLife$selfcare_score_S3)

# Just for Cognition and Social, compare first caffeine usage
GradLife_CogSoc<-GradLife%>%
  filter(area %in% c("Cognitive","Social"))%>%
  group_by(area)

ggplot(GradLife_CogSoc, aes(x=area,y=caffeine_mg_S1))+
  geom_boxplot()

```


# T-Test
```{r}
# View average self-care by area
library(dplyr)
GradLife %>%
  group_by(area) %>%
  summarise(mean_selfcare_S3 = mean(selfcare_score_S3, na.rm = TRUE),
            sd_selfcare_S3 = sd(selfcare_score_S3, na.rm = TRUE),
            n = n())

# Filter only clinical and developmental lab members
GradLife_ClinDev<-GradLife%>%
  filter(area %in% c("Clinical","Developmental"))%>%
  group_by(area)

# Check how many people per lab
table(GradLife_ClinDev$area)

# Run the t-test
t_test_result <- t.test(selfcare_score_S3 ~ area, data = GradLife_ClinDev)
print(t_test_result)
```




# Correlations
```{r}
# Plot the relationship
ggplot(GradLife, aes(x = caffeine_mg_S1, y = tiktok_min_S1)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "Caffeine T1 vs Tiktok T1")

# Correlation test
cor.test(GradLife$caffeine_mg_S1, GradLife$tiktok_min_S1)

```
# ANOVA
When you want to compare means across more than two groups, a t-test isn't enough — that's where ANOVA (Analysis of Variance) comes in.

ANOVA checks if there is any significant difference in means across the groups. It doesn't tell you which groups differ — just whether at least one group is different. You can follow up with post-hoc tests if needed.
```{r}
# Check how many areas are in the dataset
table(GradLife$area)

# Fit ANOVA model
aov_model <- aov(selfcare_score_S3 ~ area, data = GradLife)
summary(aov_model)

# Plot the means with boxplots
ggplot(GradLife, aes(x = area, y = selfcare_score_S3)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Self-Care Across Areas", y = "Self-Care Score", x = "Area")

# Let's do a post-hoc on an easier ANOVA

# Which one is actually significant?

TukeyHSD(aov_model)

```

# Linear Regression
```{r}
# Run the regression
lm_model <- lm(hours_contemplating_other_paths_S3 ~ selfcare_score_S3 + submitted_grant, data = GradLife)
summary(lm_model)
```

# Logistic Regression
```{r}
log_model<-glm(submitted_grant~tiktok_min_S2 + caffeine_mg_S2, data=GradLife,family="binomial")

summary(log_model)

# Predict probabilities
GradLife$predicted_prob <- predict(log_model, type = "response")

# Plot
ggplot(GradLife, aes(x = tiktok_min_S2, y = submitted_grant)) +
  geom_jitter(height = 0.05) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  labs(title = "Grant Submission by TikTok Minutes")

```




