---
title: "Quack Fall 2025"
author: "Alyson Wong & Nina Schoener"
date: "2025-09-29"
output: html_document
---

```{r setup}

install.packages("tidyverse")
library(tidyverse)

```

Quack Session 4 - Starter Answer Key
October 1, 2026
Alyson Wong and Nina Schoener
Adapted from Emily Rosenthal & Sierra Semko Krouse 


Today, we will learn about data visualisation! (pause for cheers from the audience). To do this, meet your new best friend GGPLOT2. Let's install this library.

```{r}

# install.packages("ggplot2")
library(ggplot2)

```

#### DATA VISUALIZATION DEMO ####

# This website has some helpful info about plotting with ggplot: https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html  

# ggplot cheat sheet: https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf  

# R-graph gallery: Help choosing and creating types of plots (NOTE: only use the ggplot section on each page) https://www.r-graph-gallery.com/index.html  

Before we start, make sure you have the fruit_vegetable_servings.csv and participant_demographics.csv loaded.

```{r}
# we're going to quickly go over again how to clean, join, and pivot these data. 

veggies <- read_csv("fruit_vegetable_servings.csv")
demo <- read_csv("participant_demographics.csv")

# cleaning the demographics data 

demo_clean <- demo %>%
  rename(FavCuisine = `Favorite Cuisine`) %>%
  mutate(Gender = as.factor(Gender), 
         Location = as.factor(Location), 
         FavCuisine = as.factor(FavCuisine))


# let's join these dataframes together

clean_data <- inner_join(veggies, demo_clean, by = "PID")

# now, let's join the two dataframes together

long_veggies <- clean_data %>%
  pivot_longer(cols = starts_with("Meal"), names_to = "Meal", values_to = "Servings") %>%
  mutate(Meal = str_remove(Meal, "Meal")) %>%
  mutate(Meal = factor(Meal, levels = c("Breakfast", "Lunch", "Dinner", "Dessert")))

```

We can look at the mean number of servings by favorite cuisine. A good way to do this is to create a bar plot. 

There are a few things to note. 
First, you have to call ggplot. 
Then, you have to tell it what dataframe you want to use.
Then, using aes(), you tell it what columns you want plotted.
Finally, you tell it what kind of plot you want. This time, we will use a bar plot.  

```{r}
cuisine_summary <- long_veggies %>%
  group_by(FavCuisine) %>%
  dplyr::summarize(mean_servings = mean(Servings, na.rm = TRUE))

# Now that we have our data loaded, let's create a simple column plot of the mean servings by favorite cuisine. What do you notice?
ggplot(cuisine_summary, aes(y = mean_servings,
                    x = FavCuisine))+
  geom_bar(stat = "identity")

```

Cool! We created our first ggplot plot. 

We can also add other things to our plot. For example, we can change the bar width on our bar plot. 
```{r}
ggplot(cuisine_summary, aes(y = mean_servings,
                    x = FavCuisine))+
  geom_bar(stat = "identity", width=.5)

#What changes?
```

Or we can add labels.
```{r}

ggplot(cuisine_summary, aes(y = mean_servings,
                    x = FavCuisine))+
  geom_bar(stat = "identity", width=.5) +
  labs(title = "Mean Fruit & Vegetable Servings by Favorite Cuisine",
       x = "FavCuisine",
       y = "Mean Servings")

```

Another thing we can do is change the color or fill of the bars based on another column in our dataset. 
We can use fill = group inside aes() to fill the bars based on that column.

```{r}

ggplot(cuisine_summary, aes(y = mean_servings,
                    x = FavCuisine, 
                    fill = FavCuisine))+
  geom_bar(stat = "identity", width=.5) +
  labs(title = "Mean Fruit & Vegetable Servings by Favorite Cuisine",
       x = "FavCuisine",
       y = "Mean Servings")

```

Let's try making a new plot using a different column. Let's use the Location colummn. 
```{r}
# first we need to create a new summary dataframe with this info grouped by location

location_summary <- long_veggies %>%
  group_by(Location) %>%
  summarize(mean_servings = mean(Servings, na.rm = TRUE))

# now we can plot veggie servings by location

ggplot(location_summary, aes(y = mean_servings,
                    x = Location))+
  geom_bar(stat = "identity")

```


We can also save any plot using ggsave! (Be careful about overwriting plots though!)
```{r}
#ggsave() by default will save the last plot generated.

ggsave("new_plot.png", width = 7, height = 4)
```

