---
title: "Summer 2026 Quack Bootcamp Session 4"
author: "Nina Schoener"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(tidylog)
```


First, we need to load in our data.
```{r}
quack_responses_clean <- read_csv("quack_responses_clean.csv")
```


Now, let's try the first example that we talked about. In this example, we're interested in whether LCD lab members like Oski more than Gopnik lab members do. 
```{r}

# View average temp by lab
quack_responses_clean %>%
  group_by(Lab) %>%
  summarise(mean_rating = mean(Bear, na.rm = TRUE),
            sd_rating = sd(Bear, na.rm = TRUE),
            n = n())

# Filter only LCD and Gopnik lab members
filtered_df <- quack_responses_clean %>%
  filter(Lab %in% c("Language and Cognitive Development Lab (PI: Mahesh Srinivasan)", "Cognitive Development and Learning Lab (PI: Alison Gopnik)"))

# Check how many people per lab
table(filtered_df$Lab)

# Run the t-test
t_test_result <- t.test(Bear ~ Lab, data = filtered_df)
print(t_test_result)

# Optional: plot the comparison
ggplot(filtered_df, aes(x = Lab, y = Bear)) +
  geom_boxplot(fill="orange") +
  labs(title = "Oski Rating by Lab (LCD vs Gopnik)",
       x = "Lab", y = "Oski Rating (1-7)")


```
When you want to compare means across more than two groups, a t-test isn't enough — that's where ANOVA (Analysis of Variance) comes in.

ANOVA checks if there is any significant difference in means across the groups. It doesn't tell you which groups differ — just whether at least one group is different. You can follow up with post-hoc tests if needed.
Our example here was whether coding knowledge varies by college year. We can also treat this as a linear regression, but we will go over that later. 

```{r}
# Check how many years are in the dataset
table(quack_responses_clean$Year)

# Fit ANOVA model
aov_model <- aov(ExperienceOther ~ Year, data = quack_responses_clean)
summary(aov_model)

# Plot the means with boxplots
ggplot(quack_responses_clean, aes(x = Year, y = ExperienceOther)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Coding Experience by Year", y = "Year", x = "Coding Knowledge (Other)")

```

In order to find the relationship between two variables without attempting to predict one from the other, we can do a Pearson's correlation. Pearson's correlation is the default, but you can change it to be a Spearman's or other kind of correlation if you need it. 
```{r}
# Plot the relationship
ggplot(quack_responses_clean, aes(x = ExperienceR, y = ExperienceOther)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(title = "R Experience vs Other Coding Experience")

# Correlation test
cor.test(quack_responses_clean$ExperienceR, quack_responses_clean$ExperienceOther, method = "pearson")

```

As an example for linear regression, we can look at whether coding knowledge varies by college year. This is the same as what we did as the example for the ANOVA, just to demonstrate that multiple analyses can apply to the same questions. 
```{r}
# Convert CollegeYear to factor and set order
quack_responses_clean$Year <- factor(quack_responses_clean$Year,
                                            levels = c("Freshman", "Sophomore", "Junior", "Senior"))

# Run the regression
lm_model <- lm(ExperienceOther ~ Year, data = quack_responses_clean)
summary(lm_model)

# Visualize the results
ggplot(quack_responses_clean, aes(x = Year, y = ExperienceOther)) +
  geom_boxplot(fill = "pink") +
  labs(title = "Coding Experience by College Year")

```

Lastly, as an example for logistic regression, we can look at whether people were more likely to think that there was a correct answer to the last survey question based on the number of pets that they have. 
```{r}
# Recode CorrectAnswer to binary
quack_responses_clean <- quack_responses_clean %>%
  mutate(CorrectAnswer = ifelse(CorrectAnswer == "Yes", 1, 0))

# Run logistic regression
log_model <- glm(CorrectAnswer ~ Pets, data = quack_responses_clean, family = "binomial")
summary(log_model)

# Plot
ggplot(quack_responses_clean, aes(x = Pets, y = CorrectAnswer)) +
  geom_jitter(height = 0.05) +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  labs(title = "Belief in Correct Answer vs Pets")

```
