---
title: "Loops & Functions in R (Psych Grad R Workshop)"
author: "Sophie Regan"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
---

# Overview

In this lesson we’ll cover:

- **For loops**: iterate over indices or elements
- **While loops**: repeat until a condition is met
- **Functions**: write reusable, testable code
- **The apply family**: `apply()`, `lapply()`, `sapply()`, `tapply()`, and `purrr::map()`

We’ll use simple examples first (strings, numbers), then psychology-flavored examples (participants, trials, reaction times).

---

# Warm-up

```{r}
# 1.a Create a vector called guest_list


# 1.b Access the second element


# 2a Create a vector phone_no with three phone numbers (as character strings)


# 2b Access the second phone number


# 3a paste0() concatenates; test it
week_days <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
paste0("Today is ", week_days[2], ".")

# paste() default separator is a space; change with 'sep'

# 3b Use paste0() with answers to 1b and 2b:

```

---

# For loops

For loops use an **index** or **iterator**. The iterator takes on each value in a sequence.

```{r}
# The : operator lets us define a range
1:3

# Print the numbers 1 through 3
for (i in 1:3) {
  print(i)
}
```

We often use **indexing** to get elements of a vector:

```{r}
guest_list <- c("Willa", "Elena", "Samira")
guest_list[1]; guest_list[2]; guest_list[3]
```

A loop lets us avoid repeating similar lines:

```{r}

```

If we **add another guest**, a hard-coded range breaks. Prefer `seq_along()` or `length()`:

```{r}
guest_list <- c(guest_list, "Sophie")

# Bad: still uses 1:3
for (i in 1:3) {
  print(guest_list[i])
}

# Better: adapts to current length


# Or loop directly over elements

```

Personalize messages using a loop:

```{r}
message <- "I’m having a Halloween Party on 10/31! The party starts at 9pm. I hope to see you there!"


# Indexing across multiple vectors in one loop:

```

Save outputs you generate inside a loop:

```{r}

```

---

## Practice: loop puzzles

**Q1.** Print this pattern:

```
1
1 2
1 2 3
1 2 3 4
1 2 3 4 5
```

```{r}

```

**Q2.** First 10 Fibonacci numbers (start 0, 1, then sum previous two):

```{r}

```

**Q3.** Multiplication table for a given number (parameterize it):

```{r}

```

**Q4.** Running total from 1 to a given number (show every intermediate sum):

```{r}
given <- 3


# Manual running sum
running <- 0

```

**Q5.** Nested loops (block × trial):

```{r}

```

**Bonus Q6.** The Hokey Pokey (nested loops + strings):

```{r}
body_parts <- c("left foot", "right foot", "head", "left hand", "right hand", "tongue")
events <- c("you put your %s in,\n", "you take your %s out,\n", "you put your %s in,\n")
rest <- "And you shake it all about.\nYou do the hokey pokey\nAnd turn yourself around—\nThat's what it's all about!\n"


```

---

# While loops

A **while loop** repeats **until a condition is no longer true**. Use while-loops when you *don’t know* in advance how many iterations you’ll need (e.g., simulate until a criterion is reached).

```{r}
# Basic pattern
i <- 1
while (i <= 5) {
  print(i)
  i <- i + 1
}
```

## Example 1: Simulate RTs until N valid trials

Imagine collecting reaction times until you have **N non-missing** valid RTs.

```{r}
set.seed(1)

```

## Example 2: Convergence (stop when change is small)

```{r}
value <- 0
delta <- Inf
iter <- 0
threshold <- 1e-4
max_iter <- 1000

```

## Safety tips for while-loops

- Always make sure **something changes** inside the loop, or you’ll get an infinite loop.
- Consider a **max iteration** safeguard.
- Prefer `for` loops when you know the number of iterations; prefer `while` when you don’t.

---

# Functions

A **function** is a named, reusable block of code. It takes **arguments**, does some work, and **returns** a value (implicitly the last expression, or explicitly with `return()`).

```{r}
# Minimal version (implicit return)


# Using a function and assigning the output
# x <- mean_rt_safe(df, col = "rt")
```

### Psychology example: per-participant summaries

```{r}
# Suppose df has columns: participant_id, rt, condition
# by_participant is a list of data frames, one per participant
# by_participant <- split(df, df$participant_id)

# Apply version (list -> numeric vector)
# avg_by_p <- sapply(by_participant, mean_rt)

# Loop version
# avg_by_p2 <- numeric(length(by_participant))
# i <- 1
# for (p in names(by_participant)) {
#   avg_by_p2[i] <- mean_rt(by_participant[[p]])
#   i <- i + 1
# }
```

---

# From loops to the *apply()* family

These are **loop alternatives** that often produce shorter, clearer, and faster code.

- `apply(X, MARGIN, FUN, ...)` — apply a function to rows/columns of a matrix/data frame (`MARGIN=1` rows, `2` cols).  
- `lapply(X, FUN, ...)` — apply a function to each element of a **list**, returns a **list**.  
- `sapply(X, FUN, ...)` — like `lapply()` but **simplifies** output to a vector/matrix if possible.  
- `tapply(X, INDEX, FUN, ...)` — apply a function to subsets defined by a **factor**.  
- `purrr::map()` — tidyverse mapping with consistent return types (e.g., `map_dbl`, `map_df`).

**Base R examples:**

```{r}
# Mean per column (matrix or data frame)
# apply(df, 2, mean, na.rm = TRUE)

# Apply to each participant's data frame in a list
# participant_means <- lapply(by_participant, mean_rt)

# Simplify to a numeric vector
# participant_means_vec <- sapply(by_participant, mean_rt)

# Mean RT by condition
# tapply(df$rt, df$condition, mean, na.rm = TRUE)
```

**Tidyverse example:**

```{r}
# library(purrr); library(readr)
# files <- c("P01.csv", "P02.csv", "P03.csv")
# results <- files %>%
#   map(readr::read_csv) %>%
#   map_dbl(~ mean(.x$rt, na.rm = TRUE))
```

---

# Exercises (hands-on)

1. **For loop**: Given `parts <- c("left", "right")` and `items <- c("shoe", "glove")`, print “left shoe”, “right shoe”, “left glove”, “right glove” using a **nested loop**.
2. **While loop**: Simulate drawing RTs (Normal(600, 80)) until you collect 20 RTs between 300 and 1200 ms. Store them in a vector and print the mean.
3. **Function**: Write `zscore_vec(x)` that returns `(x - mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)`. Then write `zscore_col(df, col)` that z-scores a column by name and returns the modified data frame.
4. **Apply**: Split a data frame by `participant_id` and compute mean RT per participant using both `sapply(split(...))` and a `for` loop. Compare the two results to confirm they match.

> Tip: try each task *both ways* (explicit loop vs. apply/map) to build fluency.
