---
title: "Loops & Functions in R (Psych Grad R Workshop)"
author: "Sophie Regan"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
---

# Overview

In this lesson we’ll cover:

- **For loops**: iterate over indices or elements
- **While loops**: repeat until a condition is met
- **Functions**: write reusable, testable code
- **The apply family**: `apply()`, `lapply()`, `sapply()`, `tapply()`, and `purrr::map()`

We’ll use simple examples first (strings, numbers), then psychology-flavored examples (participants, trials, reaction times).

---

# Warm-up

```{r}
# 1.a Create a vector called guest_list
guest_list <- c("Willa", "Elena", "Samira")

# 1.b Access the second element
guest_list[2]

# 2a Create a vector phone_no with three phone numbers (as character strings)
phone_no <- c("3459430876", "2067389942", "5482240381")

# 2b Access the second phone number
phone_no[2]

# 3a paste0() concatenates; test it
week_days <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")
paste0("Today is ", week_days[2], ".")

# paste() default separator is a space; change with 'sep'
paste("Today is", week_days[2], ".", sep = "-")

# 3b Use paste0() with answers to 1b and 2b:
paste0(guest_list[2], "'s phone number is ", phone_no[2], ".")
```

---

# For loops

For loops use an **index** or **iterator**. The iterator takes on each value in a sequence.

```{r}
# The : operator lets us define a range
1:3

# Print the numbers 1 through 3
for (i in 1:3) {
  print(i)
}
```

We often use **indexing** to get elements of a vector:

```{r}
guest_list <- c("Willa", "Elena", "Samira")
guest_list[1]; guest_list[2]; guest_list[3]
```

A loop lets us avoid repeating similar lines:

```{r}
for (i in 1:3) {
  print(guest_list[i])
}
```

If we **add another guest**, a hard-coded range breaks. Prefer `seq_along()` or `length()`:

```{r}
guest_list <- c(guest_list, "Sophie")

# Bad: still uses 1:3
for (i in 1:3) {
  print(guest_list[i])
}

# Better: adapts to current length
for (i in seq_along(guest_list)) {
  print(guest_list[i])
}

# Or loop directly over elements
for (guest in guest_list) {
  print(guest)
}
```

Personalize messages using a loop:

```{r}
message <- "I’m having a Halloween Party on 10/31! The party starts at 9pm. I hope to see you there!"

for (i in seq_along(guest_list)) {
  cat(paste0("hi ", guest_list[i], " ! ", message, "\n"))
}

# Indexing across multiple vectors in one loop:
for (i in seq_along(guest_list)) {
  cat(paste0("To: ", phone_no[i], " — hi ", guest_list[i], " — ", message, "\n"))
}
```

Save outputs you generate inside a loop:

```{r}
sent <- character(0)

for (i in seq_along(guest_list)) {
  sent[i] <- paste0("To: ", phone_no[i], " — hi ", guest_list[i], " — ", message)
}
sent

# Or append
sent2 <- character(0)
for (i in seq_along(guest_list)) {
  sent2 <- c(sent2, paste0("To: ", phone_no[i], " — hi ", guest_list[i], " — ", message))
}
sent2
```

---

## Practice: loop puzzles

**Q1.** Print this pattern:

```
1
1 2
1 2 3
1 2 3 4
1 2 3 4 5
```

```{r}
for (i in 1:5) {
  print(1:i)
}

# Alternative (build a growing vector)
pattern <- c()
for (i in 1:5) {
  pattern <- c(pattern, i)  # or: pattern[i] <- i
  print(pattern)
}
```

**Q2.** First 10 Fibonacci numbers (start 0, 1, then sum previous two):

```{r}
a <- 0; b <- 1; number_of_terms <- 10
for (i in 1:number_of_terms) {
  print(a)
  next_num <- a + b
  a <- b
  b <- next_num
}
```

**Q3.** Multiplication table for a given number (parameterize it):

```{r}
mult <- 5   # which number?
max_mult <- 10  # how far?
mult_tbl <- numeric(max_mult)
for (i in 1:max_mult) {
  mult_tbl[i] <- mult * i
}
mult_tbl
```

**Q4.** Running total from 1 to a given number (show every intermediate sum):

```{r}
given <- 3

# Using sum(1:i)
for (i in 1:given) {
  print(sum(1:i))
}

# Manual running sum
running <- 0
for (i in 1:given) {
  running <- running + i
  print(running)
}
```

**Q5.** Nested loops (block × trial):

```{r}
for (block in 1:5) {
  for (trial in 1:4) {
    print(paste0("block = ", block, " and trial = ", trial))
  }
}
```

**Bonus Q6.** The Hokey Pokey (nested loops + strings):

```{r}
body_parts <- c("left foot", "right foot", "head", "left hand", "right hand", "tongue")
events <- c("you put your %s in,\n", "you take your %s out,\n", "you put your %s in,\n")
rest <- "And you shake it all about.\nYou do the hokey pokey\nAnd turn yourself around—\nThat's what it's all about!\n"

for (part in body_parts) {
  for (ev in events) {
    cat(sprintf(ev, part))
  }
  cat(rest, "\n")
}
```

---

# While loops

A **while loop** repeats **until a condition is no longer true**. Use while-loops when you *don’t know* in advance how many iterations you’ll need (e.g., simulate until a criterion is reached).

```{r}
# Basic pattern
i <- 1
while (i <= 5) {
  print(i)
  i <- i + 1
}
```

## Example 1: Simulate RTs until N valid trials

Imagine collecting reaction times until you have **N non-missing** valid RTs.

```{r}
set.seed(1)
N <- 5
valid_rts <- numeric(0)

while (length(valid_rts) < N) {
  rt <- rnorm(1, mean = 600, sd = 80)   # draw
  is_valid <- runif(1) > 0.15           # 85% chance valid
  if (is_valid && rt > 0) {
    valid_rts <- c(valid_rts, rt)
  }
}
valid_rts
```

## Example 2: Convergence (stop when change is small)

```{r}
value <- 0
delta <- Inf
iter <- 0
threshold <- 1e-4
max_iter <- 1000

while (delta > threshold && iter < max_iter) {
  iter <- iter + 1
  new_value <- 0.5 * value + 1  # toy update rule
  delta <- abs(new_value - value)
  value <- new_value
}

list(final_value = value, iters = iter, delta = delta)
```

## Safety tips for while-loops

- Always make sure **something changes** inside the loop, or you’ll get an infinite loop.
- Consider a **max iteration** safeguard.
- Prefer `for` loops when you know the number of iterations; prefer `while` when you don’t.

---

# Functions

A **function** is a named, reusable block of code. It takes **arguments**, does some work, and **returns** a value (implicitly the last expression, or explicitly with `return()`).

```{r}
# Minimal version (implicit return)
mean_rt <- function(data, col = "rt", na_rm = TRUE) {
  mean(data[[col]], na.rm = na_rm)
}

# Safer, with checks and explicit return
mean_rt_safe <- function(data, col = "rt", na_rm = TRUE) {
  stopifnot(is.data.frame(data))
  if (!col %in% names(data)) stop(sprintf("Column '%s' not found.", col))
  x <- data[[col]]
  if (!is.numeric(x)) stop(sprintf("Column '%s' must be numeric.", col))
  return(mean(x, na.rm = na_rm))
}

# Using a function and assigning the output
# x <- mean_rt_safe(df, col = "rt")
```

### Psychology example: per-participant summaries

```{r}
# Suppose df has columns: participant_id, rt, condition
# by_participant is a list of data frames, one per participant
# by_participant <- split(df, df$participant_id)

# Apply version (list -> numeric vector)
# avg_by_p <- sapply(by_participant, mean_rt)

# Loop version
# avg_by_p2 <- numeric(length(by_participant))
# i <- 1
# for (p in names(by_participant)) {
#   avg_by_p2[i] <- mean_rt(by_participant[[p]])
#   i <- i + 1
# }
```

---

# From loops to the *apply()* family

These are **loop alternatives** that often produce shorter, clearer, and faster code.

- `apply(X, MARGIN, FUN, ...)` — apply a function to rows/columns of a matrix/data frame (`MARGIN=1` rows, `2` cols).  
- `lapply(X, FUN, ...)` — apply a function to each element of a **list**, returns a **list**.  
- `sapply(X, FUN, ...)` — like `lapply()` but **simplifies** output to a vector/matrix if possible.  
- `tapply(X, INDEX, FUN, ...)` — apply a function to subsets defined by a **factor**.  
- `purrr::map()` — tidyverse mapping with consistent return types (e.g., `map_dbl`, `map_df`).

**Base R examples:**

```{r}
# Mean per column (matrix or data frame)
# apply(df, 2, mean, na.rm = TRUE)

# Apply to each participant's data frame in a list
# participant_means <- lapply(by_participant, mean_rt)

# Simplify to a numeric vector
# participant_means_vec <- sapply(by_participant, mean_rt)

# Mean RT by condition
# tapply(df$rt, df$condition, mean, na.rm = TRUE)
```

**Tidyverse example:**

```{r}
# library(purrr); library(readr)
# files <- c("P01.csv", "P02.csv", "P03.csv")
# results <- files %>%
#   map(readr::read_csv) %>%
#   map_dbl(~ mean(.x$rt, na.rm = TRUE))
```

---

# Exercises (hands-on)

1. **For loop**: Given `parts <- c("left", "right")` and `items <- c("shoe", "glove")`, print “left shoe”, “right shoe”, “left glove”, “right glove” using a **nested loop**.
2. **While loop**: Simulate drawing RTs (Normal(600, 80)) until you collect 20 RTs between 300 and 1200 ms. Store them in a vector and print the mean.
3. **Function**: Write `zscore_vec(x)` that returns `(x - mean(x, na.rm=TRUE))/sd(x, na.rm=TRUE)`. Then write `zscore_col(df, col)` that z-scores a column by name and returns the modified data frame.
4. **Apply**: Split a data frame by `participant_id` and compute mean RT per participant using both `sapply(split(...))` and a `for` loop. Compare the two results to confirm they match.

> Tip: try each task *both ways* (explicit loop vs. apply/map) to build fluency.
