---
title: "Quack Fall 2026"
author: "Nina Schoener"
date: "2026-10-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R Fall 2026 Session 5 - Starter Code
October 8, 2026
Alyson Wong and Nina Schoener
Adapted from Emily Rosenthal & Sierra Semko Krouse 


Today, we will continue learning about data visualisation! (pause for cheers from the audience). To do this, make sure you have installed and loaded your new best friend GGPLOT2. 

```{r}
#install.packages("ggplot2")
library(ggplot2)
library(tidyverse)
```

#### DATA VISUALIZATION DEMO ####

# This website has some helpful info about plotting with ggplot: https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html  

# ggplot cheat sheet: https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf  

# R-graph gallery: Help choosing and creating types of plots (NOTE: only use the ggplot section on each page) https://www.r-graph-gallery.com/index.html  

#Today, we will be using "coffee_data.csv", which has data about participants, how much coffee they drink, and their ratings of Pumpkin Spice Lattes. 

```{r}

# Before we start, make sure you have coffee_data loaded. We're going to use a shortcut to turn all columns that are character vectors into factors (which is more or less what we want for this dataset). 

coffee_data <- read_csv("coffee_data.csv") %>%
  mutate(across(where(is.character), as.factor))

# Let's also rename some of the columns to be easier to work with

coffee_data <- coffee_data %>%
  rename("PID" = Participant_ID,
         "Pumpkin_Spice" = Pumpkin_Spice_Rating,
         "Fav_Season" = Favorite_Season,
         "Num_Cups" = Number_Cups_Coffee_Per_Week,
         "Location" = Geographic_Region)

# Now, let's create two dataframes: one that represents each person's coffee preferences and habits, and one that represents each person's favorite season

pumpkin_spice <- coffee_data %>%
  select(c("PID", "Pumpkin_Spice"))


season <- coffee_data %>%
  select(c("PID", "Fav_Season"))

# Now, let's join them together using what we learned two weeks ago about joining dataframes. 

plot_data <- left_join(pumpkin_spice, season)

# We now have a dataset that we're going to use to plot our data! 


```


```{r}

# First, let's see how what season is the most popular. From here on out, we're going to use our plot_data dataframe. 

ggplot(plot_data, aes(x = Fav_Season)) + 
  geom_histogram(stat = "count")

# What do you notice about this plot?

plot_data <- plot_data %>%
  mutate(Fav_Season = factor(Fav_Season, levels = c("Winter", "Spring", "Summer", "Fall")))

# that's right, we needed to refactor it to put the levels in the right order!

ggplot(plot_data, aes(x = Fav_Season)) + 
  geom_histogram(stat = "count")

```
Cool! We created our first ggplot plot. There are a few things to note. 
First, you have to call ggplot.
Then, you have to tell it what dataframe you want to use.
Then, using aes(), you tell it what columns you want plotted.
Finally, you tell it what kind of plot you want. This time, we used a histogram. 

We can also add other things to our plot. For example, we can add labels. By default, R will use your column names, but our y axis here for example does not have a name because we asked R to count the number of responses of each Season, i.e., that data did not come directly from a column. 

```{r}

ggplot(plot_data, aes(x = Fav_Season)) + 
  geom_histogram(stat = "count") +
  labs(
    title = "Favorite Seasons",
    x = "Season",
    y = "Number of Responses"
  )


# We may also want to have an intuitive scale on the y axis. Let's change it to 0 - 40. 

ggplot(plot_data, aes(x = Fav_Season)) + 
  geom_histogram(stat = "count") +
  labs(
    title = "Favorite Seasons",
    x = "Season",
    y = "Number of Responses"
  )+
  ylim(0, 40)

# We can also change where the breaks happen in our axes. Right now, they are set to increments of 10. Let's change it to increments of 5. 

ggplot(plot_data, aes(x = Fav_Season)) + 
  geom_histogram(stat = "count") +
  labs(
    title = "Favorite Seasons",
    x = "Season",
    y = "Number of Responses"
  )+
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 50, by = 5))

  

```

Another thing we can do is change the color or fill of the bars. 

```{r}

# We can use fill = Fav_Season inside aes() to fill the bars based on that column.

ggplot(plot_data, aes(x = Fav_Season, fill = Fav_Season)) + 
  geom_histogram(stat = "count") +
  labs(
    title = "Favorite Seasons",
    x = "Season",
    y = "Number of Responses"
  )+
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 50, by = 5))

# This also adds a legend by default. It's a bit redundant here, so we can remove it using theme().  

ggplot(plot_data, aes(x = Fav_Season, fill = Fav_Season)) + 
  geom_histogram(stat = "count") +
  labs(
    title = "Favorite Seasons",
    x = "Season",
    y = "Number of Responses"
  )+
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 50, by = 5))+
  theme(legend.position = "none")

# It will use a default color palette if you don't tell ggplot which colors to use, but you can (and should) change the colors to be whatever you want. You can do this using scale_[fill/color]_[discrete/continuous](). 

ggplot(plot_data, aes(x = Fav_Season, fill = Fav_Season)) + 
  geom_histogram(stat = "count") +
  labs(
    title = "Favorite Seasons",
    x = "Season",
    y = "Number of Responses"
  )+
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 50, by = 5))+
  theme(legend.position = "none")+
  scale_fill_discrete(palette = "Dark2")

# Go ahead and try out other palettes (you can find them on RGraphGallery linked above). For example, there's Pastel1, Accent, and Set1. 
  
  

```

The theme layer is very powerful: you can change all kinds of things about the theme you're using to plot your data inside of the theme layer. 

```{r}

# I think it's nice to center plot titles and that's one thing you can do within theme().  

ggplot(plot_data, aes(x = Fav_Season, fill = Fav_Season)) + 
  geom_histogram(stat = "count") +
  labs(
    title = "Favorite Seasons",
    x = "Season",
    y = "Number of Responses"
  )+
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 50, by = 5))+
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5))+
  scale_fill_discrete(palette = "Dark2")

# Another thing that's useful that you might want to do is rotate your x axis labels especially if your labels are very long. 

ggplot(plot_data, aes(x = Fav_Season, fill = Fav_Season)) + 
  geom_histogram(stat = "count") +
  labs(
    title = "Favorite Seasons",
    x = "Season",
    y = "Number of Responses"
  )+
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 50, by = 5))+
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10))+
  scale_fill_discrete(palette = "Dark2")


```
Before we explore more about themes, let's try out some other kinds of plots. Instead of a histogram, we can create something similar: a column plot. The difference is that we specify a column for both the x and the y axes. 
```{r}

# One other thing we might want to plot is mean pumpkin spice ratings based on favorite season, so let's calculate that and add a column for it in our data. 

plot2_data <- plot_data %>%
  group_by(Fav_Season) %>%
  summarize(MeanRating = mean(Pumpkin_Spice))


ggplot(plot2_data, aes(x = Fav_Season, y = MeanRating, fill = Fav_Season)) + 
  geom_col(position = "dodge2") +
  labs(
    title = "Pumpkin Spice Ratings by Favorite Season",
    x = "Favorite Season",
    y = "Rating"
  ) +
  scale_fill_discrete(palette = "Dark2")+
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, by = 2))+
  theme(legend.position = "none", 
        plot.title = element_text(hjust = 0.5))


```

Let's create a different type of plot. If we want to see whether number of cups of coffee per week is related to age, a scatterplot might be a good choice. 

```{r}

# the way to create a scatterplot in ggplot is using geom_point()
ggplot(coffee_data, aes(x = Age, y = Num_Cups)) + 
  geom_point()

# let's go ahead and spruce up this plot the way we did before
ggplot(coffee_data, aes(x = Age, y = Num_Cups)) + 
  geom_point(fill = "lightblue", color = "black", shape = 23, size = 3) +
  labs(
    title = "Number of Cups of Coffee Per Week by Age",
    x = "Age",
    y = "Number of Cups of Coffee"
  ) +
  scale_y_continuous(limits = c(0, 35), breaks = seq(0, 35, by = 5)) +
  scale_x_continuous(limits = c(18, 38), breaks = seq(18, 38, by = 5)) + 
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))


# In two weeks, we will explore some light stats, which will include for example adding 
# a regression line to a plot like this!


```

One thing that may be useful is saving a plot as a variable so that you can build on it later. 
```{r}

# we can also save a plot as a variable like this
plot <- ggplot(coffee_data, aes(x = Age, y = Num_Cups)) + 
  geom_point(fill = "lightblue", color = "black", shape = 23, size = 3) +
  labs(
    title = "Number of Cups of Coffee Per Week by Age",
    x = "Age",
    y = "Number of Cups of Coffee"
  ) +
  scale_y_continuous(limits = c(0, 35), breaks = seq(0, 35, by = 5)) +
  scale_x_continuous(limits = c(18, 38), breaks = seq(18, 38, by = 5)) + 
  theme(legend.position = "none", plot.title = element_text(hjust = 0.5))

# now you can call the plot using its name
plot

# or you can edit it
plot +
  theme_bw()

# try this out and try applying different themes! some options are grey, bw, linedraw, classic, minimal, void, dark, and light. the ggthemes package contains many more! themes change many things about your plots including the background, gridlines, borders, font, layout, and more. 


```


Now, let's create one last plot and put it all together! What kind of plot could make sense if we want to know if people's pumpkin spice latte rating and favorite season are based on their location?

```{r}

# lets try a few different options

# first, let's put the months in order for BirthMonth and the seasons in order for Season
coffee_data <- coffee_data %>%
  mutate(Fav_Season = factor(Fav_Season, levels = c("Winter", "Spring", "Summer", "Fall"))) %>%
  mutate(Location = factor(Location, levels = c("Northeast", "Southeast", "Midwest", "Southwest", "West")))
  
# and create a scatterplot
ggplot(coffee_data, aes(x = Location, y = Pumpkin_Spice, fill = Fav_Season)) + 
  geom_point(shape = 23, size = 3) +
  labs(
    title = "Pumpkin Spice Latte Rating & Favorite Season by Location",
    x = "Location",
    y = "Pupmkin Spice Rating", 
    fill = "Favorite Season"
  ) +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) 

# this is hard to interpret. 
# another option could be switching BirthMonth and Season, depending on which one we care more about
# we can use geom_jitter() to jitter the points slightly in case they would otherwise overlap

ggplot(coffee_data, aes(x = Location, y = Pumpkin_Spice, fill = Fav_Season)) + 
  geom_jitter(width = 0.15, size = 3) +
  labs(
    title = "Pumpkin Spice Latte Rating & Favorite Season by Location",
    x = "Location",
    y = "Pupmkin Spice Rating", 
    fill = "Favorite Season"
  ) +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) 

# this plot is still kind of hard to interpret. what if we split up the data somehow?

ggplot(coffee_data, aes(x = Location, y = Pumpkin_Spice)) + 
  facet_wrap( ~ Fav_Season) + 
  geom_jitter(width = 0.15, size = 3) +
  labs(
    title = "Pumpkin Spice Latte Rating & Favorite Season by Location",
    x = "Location",
    y = "Pupmkin Spice Rating"
  ) +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) 

# now it's a bit easier to see! It seems like people whose favorite season is Fall or Winter rate Pumpkin Spice Lattes higher than people whose favorite season is Spring or Summer. It seems like this may also not vary much by location!

# lets save this plot to a variable in case we want to use it later (we will in about 30 seconds)

facet_plot <- ggplot(coffee_data, aes(x = Location, y = Pumpkin_Spice)) + 
  facet_wrap( ~ Fav_Season) + 
  geom_jitter(width = 0.15, size = 3) +
  labs(
    title = "Pumpkin Spice Latte Rating & Favorite Season by Location",
    x = "Location",
    y = "Pupmkin Spice Rating"
  ) +
  theme(plot.title = element_text(hjust = 0.5), 
        axis.text.x = element_text(angle = 45, hjust = 1, size = 10)) 

```

We can export any plot using ggsave! (Be careful about overwriting plots though!)
```{r}

facet_plot
ggsave("facet_plot.png", width = 7, height = 4)
```

