---
title: "Quack Summer 2025"
author: "Sophie Regan"
date: "2025-07-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R Bootcamp Session 5 - Starter Code
October 6, 2025
Sophie Regan and Nina Schoener
Adapted from Emily Rosenthal & Sierra Semko Krouse


#### DATA VISUALIZATION DEMO ####

# This website has some helpful info about plotting with ggplot: https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html  

# ggplot cheat sheet: https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf  

# R-graph gallery: Help choosing and creating types of plots (NOTE: only use the ggplot section on each page) https://www.r-graph-gallery.com/index.html  

As usual, we want to make sure our libraries and data are loaded correctly. 
```{r loading libs and data}
# install.packages("ggplot2")
# install.packages("tidyverse")
library(ggplot2)
library(tidyverse)


quack_responses_clean <- read_csv("quack_responses_clean.csv")

```


Last week we practiced making a variety of bar plots. Now, let's create a scatterplot of R experience by favorite temperature. 
```{r scatterplot_numeric}

quack_responses_clean <- read_csv("quack_responses_clean.csv")


```

Let's try the same thing but with a categorical variable on the x-axis. One option could be to plot R experience grouped by whether people think a hotdog is a sandwich or not. 
```{r scatterplot_categorical}

# first, we'll create a summary dataframe of R experience grouped by hotdog opinion


# now we can create the plot



# by default our axis is labeled with a range from 2.0-4.0, but we really want it to be 1-5. let's change this now. 


# we can also change the color and size of the points on our plot to make it clearer


# we can also change the theme. there are many options to choose from, including theme_minimal(), theme_classic(), theme_gray(), theme_bw(), theme_light(), theme_dark(), and theme_void().
# you can try a few of them out and pick your favorite. 


# you will sometimes also need to change the size of all of the text in your plot at once. you can do this using base_size(). there are other ways to do this if you want to change specific elements of the text, e.g. only the title. 



# having points by themselves with no error bars is a little bit weird. let's try displaying this data using a different kind of plot: a boxplot. (in a moment, we'll learn how to add error bars to simple bar plots and scatterplots)


```


When we collect one sample, we often want to know how uncertain our estimate is. One way to do this is using bootstrapping, which is a non-parametric method that resamples your data many times to build a distribution of your statistic (e.g., the mean).

Bootstrapping is especially helpful when:

The sample size is small.

We don’t want to assume normality.

We want to estimate confidence intervals for stats that don’t have exact formulas (e.g., the median).


```{r bootstrapping}
library(boot)

# Define a function to compute the statistic of interest
boot_mean <- function(data, indices) {
  d <- data[indices, ]
  return(mean(d$FavTemp, na.rm = TRUE))
}

# Run the bootstrap
boot_results <- boot(data = quack_responses_clean, 
                     statistic = boot_mean, 
                     R = 1000)

# Print results
print(boot_results)

# Bootstrap confidence intervals
boot.ci(boot_results, type = "perc")


# Plot histogram of bootstrap sample means


```


To end, let's try out the different theme options and vote for our favorite!
We'll also try out setting the color scale using scale_fill_brewer(). 

```{r themes}

theme_names <- c("gray", "linedraw", "light", "dark", "minimal", "classic", "void")

votes <- c(0, 3, 1, 1, 4, 1, 0)

fav_themes <- as.data.frame(cbind(theme_names, votes))

# Make the plot!

```

Let's save our last plot using ggsave like we did last week. 
```{r save}

```

